Formalization of Network Flow Algorithms

### 2017-02-23 paper/root.tex
draft for preflow-push section

  \section{Push-Relabel algorithms}\label{sec:push_relabel}
  In this section we present the formalization of Push-relabel algorithms by
  Goldberg(XXX) that build up on Karzanov's idea of a preflow ---a relaxation
  of the flow definition by allowing the sum of flows over incoming edges to be
  greater equal than the sum of flows over outgoing edges of each vertex of the
  network. Karzanov was the first to leave the augmenting path idea in research
  on finding the maximum-flow in networks. He came up with an $O(V^3)$
  maximum-flow algorithm by relaxing some of the invariants in Dinitz's
  algorithm (XXX Dinitz paper). Years later, Goldberg proposed the generic
  Push-relabel algorithm which shares some ideas with Karzanov's algorithm, yet
  it provides freedom for choosing the order of ``push'' and ``relabel''
  operations. The push–relabel algorithm is considered one of the most efficient
  maximum flow algorithms. The generic algorithm has a strongly polynomial
  $O(V^2 E)$ time complexity, while specific variants of the algorithms achieve
  even lower time complexities up to $O(V^2 \sqrt{E})$. In what follows we first
  overview the transition from Ford-Fulkerson method to Push-relabel algorithm,
  and then provide the formal explanation of Goldberg's generic algorithm.

  One year after publication of Edmonds-Karp algorithm and without being aware
  of their result, Dinitz proposed another strongly polynomial algorithm based
  on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
  Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
  a smart data structure. As finding the augmenting path is the bottleneck in each
  iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
  paths from source to sink in a data-structure called the layered-network.
  Therefore, unless the length of the shortest path from source to sink has
  increased, we could use the layered-network for finding augmenting paths.
  Finding an augmenting path in the layered-network would require O(V)
  operations which is considerably more efficient than O(E) complexity of
  finding such path by searching the residual graph. Note that initializing the
  layered-network would has O(E) complexity, however this is only required
  whenever the length of the shortest path from source to sink has increased.

  Karzanov observed that at each phase of the Dinitz's algorithm it is
  sufficient to use a blocking-flow for increasing the current flow, and
  hence determining the next layered-network. He uses the term to refere to an
  augmenting flow in the current layared-network that saturates at least one edge
  on each path from source to the sink. During each phase, Karzanov's algorithm
  maintains a preflow in a layered-network. The algorithm ``pushes'' flow
  through the network to find a blocking-flow, and then ``balances'' the
  resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
  tired to eliminate balancing operation by canceling the previous goal of
  Karzanov's algorithm for constructing a feasible flow at the end of each
  phase. He proved that balancing the flow at some vertex is equivalent to
  pushing from it in future when it will appear in some layared-network at a
  greater distance from the sink. However, Cherkarsky could not come up with the
  generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
  and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
  his algorithm into phases.

  Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
  to the next level. This algorithm abandons the idea of finding a flow in each
  phase, and also abandons the idea of global phases. Instead, Goldberg idea
  maintains a preflow in the original network and pushes flow toward the sink
  along what it estimates to be shortest paths. When the algorithm terminates,
  the preflow becomes a valid flow which is also proved to be a maximum flow.

  The Push-relabels algorithms maintain a preflow as well as a height-labeling
  during the execution. A preflow in a network with source $s$ and sink $t$ is a
  function $f: V \times V \mapsto \mathbb{R}$ that satisfies the capacity
  constraint as well as a relaxed conservation constraint\footnote{every flow
    related definition, like flow's value or flow's corresponding residual
    graph can be generalized for preflows.}. That is, for all vertices except
  the source $s$ the sum of flows over all incoming edges is greater equal than
  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a vertex
  $v$ in the network is the amount by which in flow exceeds the out flow, and a
  $v$ is overflowing if $e(v) > 0$. Moreover, a height-labeling in a $s$-$t$
  network is a funtion $l: V \mapsto \mathbb{N}$ where $l(s) =
  \left\vert{V}\right\vert$, $l(t) = 0$, and for every edge $(u, v)$ of the
  residual graph we have that $l(u) \le l(v) + 1$.

@@

Goldberg distinguishes two types of pushes which are latered used for proving
the correctness and complexity of his algorithm. A push is called saturating if
the edge $(u, v)$ is removed from the residual graph after the push operation,
i.e. $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
edge $(u, v)$ remains in the residual graph after the push operation.

@@

\texttt{---Present either the refinement code or pseudocode---}

\texttt{---Discuss correctness and termination proofs if required}

\texttt{---Corelate to the FIFO and gap heuristic implementations}

\texttt{---Discuss our time complexity proof, if required}

>>>>

  \section{Preflow-Push algorithms}\label{sec:push_relabel}
  In this section we present the formalization of Preflow-push algorithms (a.k.a
  Push-relabel algorithms) by Goldberg et al.~\cite{GoTa88}  that build up on
  Karzanov's idea of a preflow~\cite{karzanov1974determination}. Karzanov was
  the first to leave the augmenting path idea in research on finding the
  maximum-flow in networks. He came up with an $O(V^3)$ maximum-flow algorithm
  by relaxing some of the invariants in Dinitz's algorithm~\cite{Di06}. Years
  later, Goldberg proposed the generic Preflow-push algorithm which shares some
  ideas with Karzanov's algorithm, yet it provides freedom for choosing the
  order of ``push'' and ``relabel'' operations. The Preflow-push algorithm is
  considered one of the most efficient maximum flow algorithms. The generic
  algorithm has a strongly polynomial $O(V^2 E)$ time complexity, while specific
  variants of the algorithms achieve even lower time complexities up to $O(V^2
  \sqrt{E})$.

  % TODO
  % @Peter: I think this first paragraph is enough for history, if you think
  % more is required, you can look into the following paragraphs (for intro mabybe)
  % TODO

  %One year after publication of Edmonds-Karp algorithm and without being aware
  %of their result, Dinitz proposed another strongly polynomial algorithm based
  %on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
  %Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
  %a smart data structure. As finding the augmenting path is the bottleneck in each
  %iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
  %paths from source to sink in a data-structure called the layered-network.
  %Therefore, unless the length of the shortest path from source to sink has
  %increased, we could use the layered-network for finding augmenting paths.
  %Finding an augmenting path in the layered-network would require O(V)
  %operations which is considerably more efficient than O(E) complexity of
  %finding such path by searching the residual graph. Note that initializing the
  %layered-network would has O(E) complexity, however this is only required
  %whenever the length of the shortest path from source to sink has increased.

  %Karzanov observed that at each phase of the Dinitz's algorithm it is
  %sufficient to use a blocking-flow for increasing the current flow, and
  %hence determining the next layered-network. He uses the term to refere to an
  %augmenting flow in the current layared-network that saturates at least one edge
  %on each path from source to the sink. During each phase, Karzanov's algorithm
  %maintains a preflow in a layered-network. The algorithm ``pushes'' flow
  %through the network to find a blocking-flow, and then ``balances'' the
  %resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
  %tired to eliminate balancing operation by canceling the previous goal of
  %Karzanov's algorithm for constructing a feasible flow at the end of each
  %phase. He proved that balancing the flow at some vertex is equivalent to
  %pushing from it in future when it will appear in some layared-network at a
  %greater distance from the sink. However, Cherkarsky could not come up with the
  %generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
  %and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
  %his algorithm into phases.

  %Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
  %to the next level. This algorithm abandons the idea of finding a flow in each
  %phase, and also abandons the idea of global phases. Instead, Goldberg idea
  %maintains a preflow in the original network and pushes flow toward the sink
  %along what it estimates to be shortest paths. When the algorithm terminates,
  %the preflow becomes a valid flow which is also proved to be a maximum flow.

  The Push-relabel algorithms maintain a preflow as well as a height-labeling
  during the execution. A preflow in a network with source $s$ and sink $t$ is a
  function $f: V \times V \mapsto \mathbb{R}$ that satisfies the capacity
  constraint as well as a relaxed conservation constraint\footnote{every flow
    related definition, like flow's value or flow's corresponding residual
    graph can be generalized for preflows.}. That is, for all vertices except
  the source $s$ the sum of flows over all incoming edges is greater equal than
  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a vertex
  $v$ in the network is the amount by which in flow exceeds the out flow, and a
  $v$ is overflowing if $e(v) > 0$. Moreover, a height-labeling in a $s$-$t$
  network is a funtion $l: V \mapsto \mathbb{N}$ where $l(s) =
  \left\vert{V}\right\vert$, $l(t) = 0$, and for every edge $(u, v)$ of the
  residual graph we have that $l(u) \le l(v) + 1$.

@@

Goldberg distinguishes two types of pushes which are later used for proving
the correctness and complexity of his algorithm. A push is called saturating if
the edge $(u, v)$ is removed from the residual graph after the push operation,
i.e. $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
edge $(u, v)$ remains in the residual graph after the push operation.

@@

\texttt{\\\\---Present either the refinement code or pseudocode---}

Correctness of the generic Push-relabel algorithm is proved in two steps. First,
we show that algorithm returns a maximum flow if it terminates. Second, we show
that the algorithm always terminates, hence, it is totaly correct.

We start with the first step, and try to prove computation of a maximum flow
upon termination of the algorithm. Maintaining helight-labeling function as well
as a preflow during the execution of the algorithm is crucial for its
correctness. Goldberg's idea is to use these functions to estimate a shortest
path for pushing flow toward the sink.

We proceed by proving that generic Push-relabel maintains a height-labeling
function. Using the definition of relabel operation $\relabel$, we observe that
the height-labeling $l(u)$ of any vertex $u \in V$ never decreases during the
execution of the algorithm, and $(l \relabel u)$ increases $u$'s height by at
least one. These two basic results prove that $l$ remains a height-labeling
whenever it is modified by applying the relabel operation. Note taht, we also
need to consider push operations. Pushing more flow along an edge $(u, v)$ in
the residual graph, not only may remove $(u, v)$ from the residual graph, but
also it can add $(v, u)$ to it. However, both of these modifications still
maintain the height-labeling function because of the preconditions that push
operator has.

The maintanance of the preflow during the execution of the algorithm can be
similiarly proved using the definition of the push operation $\push$. Next, we
proceed to prove the correctness of the algorithm, by showing that it computes a
maximum flow upon its termination. We use the fact about maintanance of
height-labeling function to build an isabelle context \isa{Labeling}. We embed
the rest of the correctness theorems inside this context, as it simplifies the
proofs and gives us a cleaner formalization. That is, as we can access the
height-labeling assumptions whenever they are required in some proof step.

As an intermediate step, we should fist prove that for any preflow $f$ and
height-labeling $l$, the residual graph corresponding to $f$ does not contain a
path from the source $s$ to the sink $t$. In order to see this, let's by
contradiction assume there is such a path. Using this path, we could obtain a
simple path $p$ conntecting $s$ to $t$. As $l$ is a valid labeling, for each
edge $(u, v)$ on $p$ we must have $l(u) \le l(v) + 1$, hence, $l(s) \le l(t) +
\vert p \vert$. This yeilds the result that $l(s) \le \vert V \vert$, since
$l(t) = 0$ and $p$ is assumed to be a simple path in the residual graph.
However, this contradicts the assumption that $l$ is a height-labeling function,
and completes the proof.

We use the previous results to complete the proof of the first step. It is easy
to see that \isai{init-preflow} and \isa{init-labeling} produce a valid preflow
as well as a valid height-labeling function. We have also shown the maintanance
of these properties earlier. When the algorithm terminates, the excess value
$e(v)$ of each vertex $v \in V$ is equal to $0$. So, the preflow that was
maintained during the execution becomes a valid flow. Given the fact that the
height-labeling assumptions are also maintained and using the latest auxillary
lemma, we can conclude that there is no path from source to sink in the residual
graph. Hence, the result of the algorithm is a maximum flow using the
Ford-fulkerson theorem.

\texttt{\\\\---Discuss termination proofs if required}

\texttt{\\\\---Discuss our time complexity proof, if required}

\texttt{\\\\---talk about relabel-to-front and fifo instances}
